---
title: "DIA Window Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DIA Window Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(MSOmicsDesign)
```

## Overview

This vignette covers the three window-design strategies provided by
**MSOmicsDesign**:

| Strategy | Function | Use case |
|----------|----------|----------|
| Fixed-width DIA | `DIA.windows()` | Equal isolation windows across the m/z range |
| Variable-width DIA | `vDIA.windows()` | Windows weighted by precursor density |
| GPF-DIA | `GPF.windows()` | Gas-phase fractionation from an existing isolation list |

All functions return an `mzdf` data frame that can be visualised with `plot()`
and saved with `save.csv = TRUE`.

---

## 1. Fixed-width DIA windows — `DIA.windows()`

### Specifying window width

```{r}
dia <- DIA.windows(mnc = 400, mxc = 1000, windows.width = 20)
head(dia)
nrow(dia) # number of windows
```

### Specifying number of windows

Use `n.windows` instead of `windows.width` to request an approximate window
count. The actual width is computed as
`ceiling((mxc - mnc) / n.windows)`, so the final count may be ±1.

```{r}
dia_n <- DIA.windows(mnc = 400, mxc = 1000, n.windows = 30)
nrow(dia_n)
```

### Output types

`output.type` controls the columns returned:

```{r}
# Default: lower/upper bounds
DIA.windows(mnc = 400, mxc = 1000, windows.width = 20) |> head(3)

# Center mass + scan width (for instruments that use this format)
DIA.windows(
  mnc = 400,
  mxc = 1000,
  windows.width = 20,
  output.type = "center"
) |>
  head(3)

# Both together
DIA.windows(mnc = 400, mxc = 1000, windows.width = 20, output.type = "both") |>
  head(3)
```

### Staggered windows

Staggered acquisition doubles the number of windows by offsetting a second set
by half a window width. This improves coverage at window edges.

```{r}
dia_stagger <- DIA.windows(
  mnc = 400,
  mxc = 1000,
  windows.width = 20,
  stagger = TRUE
)
nrow(dia_stagger) # 2x the non-staggered count
```

### Window optimisation (`opt.windows`)

By default `optimize = TRUE`, which places window edges in m/z "forbidden
zones" where +2H and +3H peptides are rarely present, minimising chimeric
spectra. This calls `opt.windows()` internally.

```{r}
# Manual use of opt.windows
edges <- seq(400, 1000, 20)
opt.windows(edges) |> head()

# With phospho correction
opt.windows(edges, phospho = TRUE) |> head()
```

### Visualise

```{r, fig.height = 5}
plot(dia)
```

---

## 2. Variable-width DIA windows — `vDIA.windows()`

Variable windows allocate narrower ranges in dense precursor regions and wider
ranges where precursors are sparse, giving roughly equal ion current per window.

The function uses the `precursor_hela` reference dataset (HeLa precursor m/z
frequencies from Webel et al. 2024) by default. You can supply your own
reference with the `ref.data` argument.

```{r}
vdia <- vDIA.windows(mnc = 400, mxc = 1000, n.windows = 32)
head(vdia)
nrow(vdia)
```

Compare window widths — narrow in the dense 400–700 region, wider elsewhere:

```{r}
vdia$width <- vdia$upper_windows - vdia$lower_windows
summary(vdia$width)
```

```{r, fig.height = 5}
plot(vdia)
```

### Custom reference data

Supply any data frame with integer `mz` and frequency `n` columns:

```{r, eval = FALSE}
my_ref <- data.frame(mz = 400:999, n = rpois(600, lambda = 100))
vDIA.windows(mnc = 400, mxc = 1000, n.windows = 32, ref.data = my_ref)
```

---

## 3. GPF-DIA windows — `GPF.windows()`

Gas-phase fractionation (GPF) runs a full DIA isolation list in multiple
injections, each covering a sub-range of the m/z axis. `GPF.windows()` takes
an existing isolation list and partitions it into groups.

### From an isolation list string vector

```{r}
data(staggered_mz)
head(staggered_mz)

gpf <- GPF.windows(staggered_mz$range, mnc = 400, mxc = 1000)
length(gpf) # number of GPF groups
names(gpf) # m/z range of each group
```

Each element is an `mzdf` data frame for one GPF injection:

```{r}
gpf[[1]]
```

### From separate lower/upper bound vectors

```{r}
mz_list <- strsplit(staggered_mz$range, split = "-")
lower_b <- as.numeric(sapply(mz_list, "[[", 1))
upper_b <- as.numeric(sapply(mz_list, "[[", 2))

gpf2 <- GPF.windows(
  lower_windows = lower_b,
  upper_windows = upper_b,
  mnc = 400,
  mxc = 1000
)
identical(names(gpf), names(gpf2))
```

### Custom GPF scan range

`windows_width` (default 100 Da) sets the m/z span of each GPF injection:

```{r}
gpf_wide <- GPF.windows(
  staggered_mz$range,
  mnc = 400,
  mxc = 1000,
  windows_width = 200
)
length(gpf_wide) # fewer groups than the default 100 Da split
```

### Visualise a single GPF group

```{r, fig.height = 3}
plot(gpf[[1]])
```

---

## Saving output

All three functions accept `save.csv = TRUE` to write their output to a `.csv`
file importable by most instrument control software:

```{r, eval = FALSE}
DIA.windows(mnc = 400, mxc = 1000, windows.width = 20, save.csv = TRUE) # writes DIA_windows.csv

vDIA.windows(mnc = 400, mxc = 1000, n.windows = 32, save.csv = TRUE) # writes variable_windows.csv
```

`GPF.windows()` writes one `.csv` per group:

```{r, eval = FALSE}
GPF.windows(staggered_mz$range, mnc = 400, mxc = 1000, save.csv = TRUE) # writes 400-500.csv, 500-600.csv, …
```

---

## Choosing a strategy

| Question | Recommendation |
|----------|---------------|
| First experiment, unknown sample complexity | Fixed-width (`DIA.windows`) |
| HeLa-like proteome, maximise identifications | Variable-width (`vDIA.windows`) |
| Deep proteome with limited MS time | GPF-DIA (`GPF.windows`) |
| Edge-chimera artefacts visible in data | Enable `optimize = TRUE` (default) |
| Low-abundance phosphopeptides | `opt.windows(phospho = TRUE)` |
