---
title: "Plate Design Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plate Design Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(MSOmicsDesign)
```

## Overview

This vignette walks through a complete plate design workflow for an MS-based
omics experiment:

1. Define the sample list
2. Insert QC samples
3. (Optional) Block-randomise by experimental covariates
4. Fill samples into plate matrices
5. Merge plates from multiple batches
6. Visualise plates
7. Export plate positions for an instrument sequence file

---

## 1. Define samples

Start with a character vector of sample identifiers. Here we use 20 samples
from a two-condition experiment.

```{r}
samples <- paste0("S", 1:20)
samples
```

---

## 2. Insert QC samples with `addQC()`

`addQC()` inserts QC runs at a fixed interval throughout the sequence.

```{r}
samples_qc <- addQC(samples, interval = 6)
samples_qc
```

A QC is placed after every 6 samples. The QCs are automatically labelled
`QC_1`, `QC_2`, … You can change the prefix with the `label` argument.

```{r}
addQC(samples, interval = 6, label = "POOL")
```

To insert QCs at specific positions instead of a regular interval, use the
`position` argument:

```{r}
addQC(samples, position = c(1, 10, 20))
```

Use `replace = TRUE` to swap samples at the given positions in-place (sequence
length unchanged) rather than inserting additional entries:

```{r}
addQC(samples, position = c(1, 10, 20), replace = TRUE)
```

Use `permute = TRUE` with `interval` to randomise the order within each block
(samples + QC shuffled together):

```{r}
addQC(samples, interval = 6, permute = TRUE)
```

---

## 3. Block randomisation with `block.rand()` (optional)

When your experiment has multiple covariates (e.g. time-point and treatment),
block randomisation ensures each run-order block contains a balanced mix.

```{r}
data(ccc)
head(ccc)
```

```{r, message = FALSE}
ccc_block <- block.rand(ccc, time, response)
ccc_block
```

The result has `block.id`, `block.size`, `unique_group`, and `unique_group_id`
columns appended to the original data frame.

### Controlling block size

By default `block.rand()` maximises the number of blocks (one sample per group
per block). Use `block_size` to set how many samples per group appear in each
block, or `n_blocks` to fix the number of blocks directly.

```{r, message = FALSE}
# 2 samples per group per block
block.rand(ccc, time, response, block_size = 2)
```

```{r, message = FALSE}
# Exactly 3 blocks
block.rand(ccc, time, response, n_blocks = 3)
```

Use `seed` to make results reproducible (default `seed = 1`).

### Checking block balance

`check.block()` tabulates each covariate combination within every block so you
can verify balance at a glance.

```{r}
check.block(ccc_block)
```

### Using block-randomised samples downstream

The `unique_group_id` column gives a unique, randomised label per sample that
can be piped directly into `fill.plate()`.

```{r}
fp_block <- fill.plate(ccc_block$unique_group_id)
```

---

## 4. Fill samples into plates with `fill.plate()`

`fill.plate()` distributes a sample vector across one or more 96-well (or
custom-size) plates.

```{r}
fp <- fill.plate(samples_qc)
class(fp)
length(fp$plate) # number of plates used
```

By default the plate is 6 rows × 9 columns (54 wells). Samples spill onto the
next plate automatically when they exceed the well count.

```{r}
# Two full plates of 60 samples
fp2 <- fill.plate(paste0("S", 1:60))
length(fp2$plate)
```

### Custom plate dimensions and names

```{r}
fill.plate(
  samples_qc,
  nrows = 8,
  ncols = 12,
  plate_names = c("Plate1", "Plate2")
)$plate |>
  names()
```

### Attaching a labels vector

A companion label vector (e.g. sample descriptions or weights) can be stored
alongside the plate matrix and displayed on plots.

```{r}
labels <- paste0("lbl_", seq_along(samples_qc))
fp_lbl <- fill.plate(samples_qc, labels = labels)
names(fp_lbl) # $plate, $label, $data
class(fp_lbl$label) # inherits "plate" — same structure as $plate
```

### Reserving wells with `exclude`

Use `exclude` to mark instrument QC wells or other reserved positions. Reserved
wells are filled with `exclude_label` (default `"iQC"`) and skipped when
placing samples.

Apply the same exclusion to every plate with a character vector:

```{r}
fill.plate(paste0("S", 1:20), exclude = "A1")$plate
```

Apply per-plate exclusions with a named list. The `"all"` key applies to every
plate; other keys apply to named plates only:

```{r}
fill.plate(paste0("S", 1:100), exclude = list(all = "A1", G = "B3"))$plate |>
  names()
```

---

## 5. Merge plates from multiple batches with `merge_plate()`

When an experiment is designed in separate batches — for example two runs
designed independently — `merge_plate()` combines two `plate_list` objects
into one.

```{r}
batch_a <- fill.plate(paste0("S", 1:20))
batch_b <- fill.plate(paste0("S", 21:40))

merged <- merge_plate(batch_a, batch_b)
length(merged$plate) # plates from both batches
names(merged$plate) # plate names
```

### Name conflicts

If both batches produce plates with the same name (e.g. both produce an `"R"`
plate), `merge_plate()` appends suffixes to disambiguate:

```{r}
big_a <- fill.plate(paste0("S", 1:60)) # fills R and G
big_b <- fill.plate(paste0("S", 61:120)) # also fills R and G

merged2 <- merge_plate(big_a, big_b)
names(merged2$plate) # "R_1" "G_1" "R_2" "G_2"
```

Use the `suffixes` argument to customise the appended strings:

```{r}
merged3 <- merge_plate(big_a, big_b, suffixes = c("_batchA", "_batchB"))
names(merged3$plate)
```

### Merging with labels

Labels are merged when both inputs have them. A warning is issued and the label
slot is dropped if only one side carries labels.

```{r}
a_lab <- fill.plate(paste0("S", 1:20), labels = paste0("L", 1:20))
b_lab <- fill.plate(paste0("S", 21:40), labels = paste0("L", 21:40))

merged_lab <- merge_plate(a_lab, b_lab)
length(merged_lab$label) # combined label plates
```

---

## 6. Visualise plates with `plot()`

`plot()` dispatches to `plot.plate_list()` for a `plate_list` object, or to
`plot.plate()` for a single plate matrix.

### No labels — colour by group

```{r, fig.height = 3}
plot(fp)
```

### Flat vector labels

Pass a vector of the same length as the original sample vector. Labels are
split across plates automatically.

```{r, fig.height = 3}
plot(fp, labels = paste0("lbl_", seq_along(samples_qc)))
```

### Plate-structure labels from `fill.plate()$label`

When labels were stored via `fill.plate(labels = ...)`, pass `$label` directly.
This is the recommended path — no splitting or padding occurs.

```{r, fig.height = 3}
plot(fp_lbl, labels = fp_lbl$label)
```

### Stripping QC suffixes for colour grouping

Use `control.suffix` to collapse `QC_1`, `QC_2`, … into a single `"QC"` colour
group on the plot.

```{r, fig.height = 3}
plot(fp, control.suffix = "QC")
```

### Plotting a single plate matrix directly

```{r, fig.height = 3}
plot(fp$plate[[1]])
```

---

## 7. Generate instrument sequence positions with `generate.position()`

`generate.position()` converts the plate matrices into position strings in the
format expected by Thermo Xcalibur (`plate_name:rowcol`, e.g. `"R:A1"`).

```{r}
pos <- generate.position(fp)
pos
```

Set `flatten = TRUE` to collapse all plates into a single character vector
ready for insertion into a sequence table.

```{r}
pos_flat <- generate.position(fp, flatten = TRUE)
head(pos_flat)
length(pos_flat) # equals length(samples_qc)
```

---

## Full workflow in one block

```{r, eval = FALSE}
library(MSOmicsDesign)

# 1. Samples
samples <- paste0("S", 1:50)

# 2. QC every 8 samples
samples_qc <- addQC(samples, interval = 8)

# 3. Fill plates (6 x 9 default), reserve A1 as instrument QC on every plate
fp <- fill.plate(
  samples_qc,
  labels = paste0("lbl_", seq_along(samples_qc)),
  exclude = "A1"
)

# 4. Merge a second batch
fp2 <- fill.plate(
  paste0("S", 51:100),
  labels = paste0("lbl_", 51:100),
  exclude = "A1"
)
fp_all <- merge_plate(fp, fp2)

# 5. Visualise
plot(fp_all, labels = fp_all$label, control.suffix = "QC")

# 6. Export positions
positions <- generate.position(fp_all, flatten = TRUE)
```
